Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-07-06 workflow/workflow.Rmd
Changes to text and figures

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages. Throughout the workflow, we use a unique `SummarizedExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.

@@

```{r, echo = FALSE}
urls = c("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz",
         "https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt")

if(!file.exists("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")) {
  download.file(urls[1], "../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
  R.utils::gunzip("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
}

if(!file.exists("../data/oeHBCdiff_clusterLabels.txt")) {
  download.file(urls[2], "../data/oeHBCdiff_clusterLabels.txt")
}
```

```{r}
load("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")

@@

# Extract metadata
batch <- droplevels(pData(Cufflinks_eSet)$MD_c1_run_id)
bio <- droplevels(pData(Cufflinks_eSet)$MD_expt_condition)
clusterLabels <- read.table("../data/oeHBCdiff_clusterLabels.txt",
                            sep = "\t", stringsAsFactors = FALSE)
m <- match(colnames(E), clusterLabels[, 1])

@@

The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figure \@ref(fig:addPublishedClusters)), the main difference being that several of the published clusters were merged here into single clusters. This discrepancy is likely caused by the fact that we performed dimensionality reduction using ZINB-WaVE and provided the resulting `W` matrix as input to `RSEC` instead of performing dimensionality reduction on the normalized counts within the `RSEC` function.

[FP: Elizabeth or Davide, how can we change the colors in Figure \@ref(fig:addPublishedClusters) to match the colors we use for publishedClusters (that is `col_clus` like in Figure \@ref(fig:mdsW)). I tried colPalette argument, but it does not work with SummarizedExperiment input. Also, would it be possible to have the same order and labels for the barplots in Figures \@ref(fig:barplotOurs) and \@ref(fig:addPublishedClusters) (barplots labelled -1 and Not Assigned are the same, right?)?]

@@

```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, "color"]
names(pal) <- palDF[, "name"]
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20,
     xlab = "Component1", ylab = "Component2")
legend(x = "bottomright", legend = names(pal), cex = .5,
       fill = pal, title = "Sample")
```

@@

Cells in cluster `c4` have a cluster label of `-2` in the original published clustering, meaning that they were not assigned to any cluster. These cells were actually identified as non-sensory contaminants, as they overexpress gene `Reg3g` (see Figure S1 from [@Fletcher2017] and Figure \@ref(fig:boxplotReg3g)), and were removed from the original published clustering. While it is reassuring that our workflow clustered these cells separately, with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes, as cells in this cluster do not participate in the cell differentiation process. Note that our workflow's clustering only captured XXX out of XXX cells overexpressing gene `Reg3g` in cluster `c4` (see Figure \@ref(fig:boxplotReg3g)), however we retain the remaining cells to infer lineages as they did not seem to influence the clustering.

[FP: check writing + table with correspondance of clusters. SD: I did minor edits to text above. Please check. Perhaps mention total number of contaminant cells and how many are in cluster `c4`.]

```{r boxplotReg3g, fig.width=4, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g for cells in cluster c4 (left) and cells not in cluster c4 (right)."}
c4 <- rep("other clusters", ncol(se)) 
c4[primaryClusterNamed(ceObj) == "c4"] <- "cluster c4"
boxplot(log1p(assay(se)["Reg3g", ]) ~ c4, ylab = "log counts", cex.axis = .8, cex.lab = .8)
```

@@

Finally, a new R class called `SingleCellExperiment` should be released soon. This new class would essentially be a `SummarizedExperiment` class with a couple of additional slots, the most important of which is `reducedDims`, which, much like the `assays` slot of `SummarizedExperiment`, can contain one or more matrices of reduced dimension. This new `SingleCellExperiment` class would be a valuable addition to the workflow, as we could store in a single object the raw counts as well as the low-dimensional matrix created by the ZINB-WaVE dimensionality reduction step. Once this class is implemented, we would like to incorporate it to the workflow.

@@

This workflow provides a tutorial for the analysis of scRNA-seq data in R. It covers four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis along lineages. The workflow is general and flexible, allowing the user to sustitute the statistical method used in each step by a different method. We hope our proposed workflow will ease technical aspects of scRNA-seq data analysis and help with the discovery of novel biological insights.

@@

The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow. The four packages used in the workflow (`scone`, `zinbwave`, `clusterExperiment`, and `slingshot`) are Bioconductor R packages and are available on GitHub at, respectively, https://github.com/YosefLab/scone, https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot.

>>>>

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages. Throughout the workflow, we use a single `SummarizedExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.

@@

Note that in the following, we assume that the user has access to a data folder located at `../data`. Users with a different directory stracture may need to change the `data_dir` variable below to reproduce the workflow.

```{r}
data_dir <- "../data/"

urls = c("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz",
         "https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt")

if(!file.exists(paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda"))) {
  download.file(urls[1], paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz"))
  R.utils::gunzip(paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz"))
}

if(!file.exists(paste0(data_dir, "oeHBCdiff_clusterLabels.txt"))) {
  download.file(urls[2], paste0(data_dir, "oeHBCdiff_clusterLabels.txt"))
}
```

```{r}
load(paste0(data_dir, "GSE95601_oeHBCdiff_Cufflinks_eSet.Rda"))

@@

# Extract metadata
batch <- droplevels(pData(Cufflinks_eSet)$MD_c1_run_id)
bio <- droplevels(pData(Cufflinks_eSet)$MD_expt_condition)
clusterLabels <- read.table(paste0(data_dir, "oeHBCdiff_clusterLabels.txt"),
                            sep = "\t", stringsAsFactors = FALSE)
m <- match(colnames(E), clusterLabels[, 1])

@@

The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figure \@ref(fig:addPublishedClusters)), the main difference being that several of the published clusters were merged here into single clusters. This discrepancy is likely caused by the fact that we started with the top 1,000 genes, which might not be enough to discriminate between closely related clusters.

@@

```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, "color"]
names(pal) <- palDF[, "name"]
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20,
     xlab = "Component1", ylab = "Component2")
legend(x = "topleft", legend = names(pal), cex = .5,
       fill = pal, title = "Sample")
```

@@

Cells in cluster `c4` have a cluster label of `-2` in the original published clustering, meaning that they were not assigned to any cluster. These cells were actually identified as non-sensory contaminants, as they overexpress gene `Reg3g` (see Figure S1 from [@Fletcher2017] and Figure \@ref(fig:boxplotReg3g)), and were removed from the original published clustering. While it is reassuring that our workflow clustered these cells separately, with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes, as cells in this cluster do not participate in the cell differentiation process. Note that, out of the 77 cells overexpressing `Reg3g`, 11 are captured in cluster `c4` and 21 are unclustered in our workflow's clustering (see Figure \@ref(fig:boxplotReg3g)). However, we retain the remaining 45 cells to infer lineages as they did not seem to influence the clustering.

```{r boxplotReg3g, fig.width=6, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g stratified by cluster."}
c4 <- rep("other clusters", ncol(se)) 
c4[primaryClusterNamed(ceObj) == "c4"] <- "cluster c4"
boxplot(log1p(assay(se)["Reg3g", ]) ~ primaryClusterNamed(ceObj), 
        ylab = "Reg3g log counts", cex.axis = .8, cex.lab = .8)
```

@@

Finally, a new S4 class called `SingleCellExperiment` is currently under development (https://github.com/drisso/SingleCellExperiment). This new class is essentially a `SummarizedExperiment` class with a couple of additional slots, the most important of which is `reducedDims`, which, much like the `assays` slot of `SummarizedExperiment`, can contain one or more matrices of reduced dimension. This new `SingleCellExperiment` class would be a valuable addition to the workflow, as we could store in a single object the raw counts as well as the low-dimensional matrix created by the ZINB-WaVE dimensionality reduction step. Once the implementation of this class is stable, we would like to incorporate it to the workflow.

@@

This workflow provides a tutorial for the analysis of scRNA-seq data in R/Bioconductor. It covers four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis along lineages. The workflow is general and flexible, allowing the user to sustitute the statistical method used in each step by a different method. We hope our proposed workflow will ease technical aspects of scRNA-seq data analysis and help with the discovery of novel biological insights.

@@

The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow. The four packages used in the workflow (`scone`, `zinbwave`, `clusterExperiment`, and `slingshot`) are Bioconductor R packages and are available at, respectively, https://bioconductor.org/packages/scone, https://bioconductor.org/packages/zinbwave, https://bioconductor.org/packages/clusterExperiment, and https://github.com/kstreet13/slingshot.
